---
title: "Paper Airplane Experiment: RCBD Design"
author: "Angela Kang"
date: "2024-11-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(knitr)
library(stats)
```

# Introduction

The aerodynamic performance of paper airplanes depends on various factors, including design and weight distribution. Proper weight placement ensures a balanced center of gravity, which is critical for stable and efficient flight. This principle applies broadly, from commercial airplanes, where fuel and cargo distribution are optimized for stability, to UAVs, where precise payload positioning impacts maneuverability and duration (Scully et al., 2019). The simplicity of paper airplanes provides an effective model to study these concepts, offering an accessible way to visualize the impact of weight displacement on flight mechanics.

Smith and Johnson (2020) further emphasize that the placement of additional mass significantly impacts flight performance in paper airplanes, highlighting the relationship between balanced weight distribution and increased flight distance. Their study revealed that strategic mass allocation can stabilize flight paths and optimize aerodynamic efficiency. Similarly, Davis et al. (2018) demonstrate that alterations to the center of gravity can either enhance or impede aerodynamic efficiency, illustrating the importance of careful weight allocation in flight mechanics. Their findings underscore how imbalanced weight distribution can destabilize aircraft, leading to compromised performance and reduced flight distances.

This study explores the effect of paperclip placement (nose, middle, and rear) on flight distance, simulating real-world scenarios in aviation. By systematically varying the placement of weights and analyzing the statistically significant results, this research highlights the aerodynamic implications of weight displacement. The middle placement, for instance, may mimic the balanced center of gravity in well-engineered aircraft, while extreme placements at the nose or rear could illustrate destabilizing effects observed in improperly balanced systems. This experiment aims to deepen our understanding of how weight distribution influences not only paper airplane performance but also broader applications in aviation engineering.

This experiment utilizes a Randomized Complete Block Design (RCBD):

- **Blocks:** Each airplane represents a block.
- **Treatments:** Four levels of paperclip placement:
  - Control (no paperclip)
  - Paperclip on the nose
  - Paperclip on the middle
  - Paperclip on the rear
- **Outcome Variable:** Flight distance (inches).

By systematically varying paperclip placement and analyzing the results, this study aims to provide insights into the interplay between weight distribution and aerodynamic performance.

# Methods

The experiment was conducted indoors to minimize external factors like wind. The 30 airplanes were folded using the same design to ensure consistency across blocks. Each airplane was tested under all four treatments in a fully randomized order.

## Experimental Setup:

- **Environment:** The experiment was conducted indoors in a large, well-lit room with minimal airflow to reduce environmental variability. This ensured consistent conditions across all trials.
- **Airplane Design:** All 30 airplanes were folded using the same step-by-step instructions to ensure uniformity. The paper used was standard 8.5 x 11-inch printer paper.
- **Treatment Levels:**
  - **Control:** No paperclip was added, providing a baseline measurement of the airplane's natural flight capability.
  - **Nose:** A single paperclip was attached at the nose of the airplane to simulate a forward weight bias.
  - **Middle:** A paperclip was attached at the midpoint of the airplane's body, representing a balanced weight distribution.
  - **Rear:** A paperclip was attached at the tail end to create a rear-heavy design.
- **Measurement Tool:** Flight distances were measured using a tape measure accurate to the nearest tenth of an inch. Measurements were taken from the starting point to the airplane's nose.
- **Throwing Technique:** To minimize variability, the same individual performed all throws. Each throw was executed with a consistent arm motion and release angle, standardized through practice trials.

## Randomization

Each airplane will undergo all four treatment conditions in a random order to reduce bias. The randomized order is done by code.

```{r, echo=TRUE}
treatments <- c("Control", "Nose", "Middle", "Rear")
airplanes <- paste("Airplane", 1:30)  # Adjust to match the number of airplanes used

design <- expand.grid(Airplane = airplanes, Treatment = treatments)

set.seed(2024)  
randomized_design <- design %>%
  sample_n(nrow(design)) %>%  
  mutate(ThrowOrder = row_number())  
```
## Statistical Methods
- ANOVA: Used to analyze the effect of treatment while accounting for block effects (airplane variability).
- Post-hoc Analysis: Tukey HSD for pairwise comparisons.
- Assumptions:
  - Normality (checked via residual diagnostics).
  - Homoscedasticity (checked via plots of residuals vs. fitted values).
  - Independent observations

# Results

## Sample Size Calculation
```{r, echo=FALSE}
k <- 4                  # Number of treatment groups
alpha <- 0.05           # Significance level
power <- 0.8            # Desired power

# Variance from preliminary experiment (used for power calculation)
sd <- 92.01177           # Standard deviation from preliminary experiment
within.var <- sd^2       # Calculated within variance

# Varying between-group variances
between.var <- seq(50, 200, by=10)

# Initialize variables for sample sizes
n_var1 <- NA
for(i in 1:length(between.var)) {
  n_var1[i] <- power.anova.test(groups = k, 
                                between.var = between.var[i], 
                                within.var = within.var, 
                                power = power, 
                                sig.level = alpha)$n
}

# Combine results into a data frame
sample_sizes <- data.frame(
  n = n_var1,
  between.var = between.var,
  within.var = rep(within.var, length(n_var1))
)

# Plot the sample size
library(ggplot2)
ggplot(data = sample_sizes, mapping = aes(x = between.var, y = n)) +
  geom_point() + geom_line() +
  labs(title = "Sample Size vs Between-Group Variance", 
       x = "Between-Group Variance", 
       y = "Sample Size") +
  theme_minimal()
```
At a lower between-group variance (e.g., around 50), the required sample size is approximately 620 blocks per group. This large sample size demonstrates the difficulty of detecting differences between groups when the variance is small. As the between-group variance increases (e.g., to 200), the required sample size decreases significantly, to around 150 blocks per group.

However, due to practical constraints and the unknown between group variance, this experiment will use 30 blocks, with each treatment applied to every group. While limiting the sample size to 30 may reduce the statistical power and increase the risk of a Type II error, the adequacy of 30 blocks per treatment can only be confirmed after data collection and analysis of the between-group variance.

## Data Collection
```{r, echo=FALSE}
# Create a data frame for recording results
results <- randomized_design %>%
  mutate(Distance = c(100.5, 159, 100.5, 107.8, 119.4, 81.2, 103.3, 110.8, 133.7, 93.2, 
                      119.5, 142.9, 94.8, 109.6, 94.4, 162, 108.9, 106.1, 110.8, 125.4, 
                      103.2, 131.4, 113.2, 127, 99.9, 165, 105.1, 105.6, 164.2, 156.6,
                      97.4, 136.8, 124.4, 153.7, 124.1, 163.1, 164, 151, 144.3, 129.7, 
                      155.5, 180.5, 151.3, 101, 101.4, 153.3, 147, 126.6, 146, 147,
                      183.2, 160, 173.4, 127, 130.7, 100.6, 127.6, 160, 90.9, 98.5, 
                      89.4, 163.2, 157, 159.8, 144.7, 135.7, 127, 153.1, 157, 155.6, 
                      166.5, 89.6, 81, 200, 87, 138.2, 126.2, 107.5, 155.7, 144.7,
                      161.2, 135.9, 196.4, 185.7, 160, 83.4, 95.5, 96.2, 160.6, 136.6, 
                      131, 193.2, 139.3, 117, 110.8, 150, 150.3, 166.2, 94.8, 157.7,
                      117.7, 186.2, 141.4, 76.9, 130.3, 145.5, 120.7, 134.2, 155, 92.7,
                      80, 117.7, 90, 96.4, 130.7, 146.1, 143.4, 124, 123.1, 86.5)) 
```

## Summary statistics
```{r, echo=FALSE}
summary_stats <- results %>%
  group_by(Treatment) %>%
  summarize(
    Mean = mean(Distance, na.rm = TRUE),
    SD = sd(Distance, na.rm = TRUE),
    N = n()
  )

kable(summary_stats, caption = "Summary Statistics by Treatment")
```

## Visualization
```{r, echo=FALSE}
# Boxplot of distances by treatment
ggplot(results, aes(x = Treatment, y = Distance)) +
  geom_boxplot() +
  labs(title = "Flight Distance by Treatment", x = "Treatment", y = "Distance (in)") +
  theme_minimal()
```

We can observe two outliers but there are only two, so they are insignificant enough to ignore. 

The summary statistics and boxplot indicate clear trends in performance across treatments. The middle placement produced the highest mean flight distance, underscoring the importance of balanced weight distribution. The rear placement consistently underperformed, with the lowest mean distance, confirming the destabilizing effects of rear-heavy configurations. The nose placement showed only slight improvement over the control, suggesting that a forward-shifted center of gravity alone does not enhance performance significantly.

### Sample Size Adequacy Test
```{r, echo=TRUE}
# Calculate the overall mean
overall_mean <- mean(results$Distance)

# Step 3: Compute the between-group sum of squares
summary_stats <- summary_stats %>%
  mutate(WeightedSqDev = N * (Mean - overall_mean)^2)

between_group_ss <- sum(summary_stats$WeightedSqDev)

# Step 4: Calculate between-group variance
k <- n_distinct(results$Treatment)  # Number of groups
between_group_variance <- between_group_ss / (k - 1)

k <- 4                  # Number of treatment groups
alpha <- 0.05           # Significance level
power <- 0.8            # Desired power
sd <- 92.01177          # Standard deviation from preliminary experiment
within.var <- sd^2      # Calculated within variance

# Calculate the sample size for the observed between-group variance
sample_size <- power.anova.test(
  groups = k, 
  between.var = between_group_variance, 
  within.var = within.var, 
  power = power, 
  sig.level = alpha
)$n
```
Based on the observed between-group variance of `r round(between_group_variance, 2)`, the required sample size per group to achieve 80% power at a significance level of 0.05 is `r ceiling(sample_size)`. Since the experiment used 30 blocks (30 observations per treatment), the sample size is adequate to achieve the desired statistical power.


## ANOVA 
```{r, echo=FALSE}
# ANOVA model
model <- aov(Distance ~ Treatment + Airplane, data = results)
anova_results <- summary(model)

kable(anova_results[[1]], caption = "ANOVA Results for RCBD")

```

The ANOVA results reveal significant differences in flight distances among the four treatment groups (Control, Nose, Middle, Rear), with a p-value < 0.001. This confirms that the placement of the paperclip has a measurable impact on the aerodynamic performance of the paper airplanes while accounting for variability between airplanes as blocks. The airplane block effect was non-significant, indicating that variability between airplanes did not overshadow treatment effects.

## Post-hoc Analysis
```{r, echo=FALSE}
# Tukey HSD
tukey_results <- TukeyHSD(model, "Treatment")
kable(as.data.frame(tukey_results$Treatment), caption = "Tukey HSD Results")
```
The experiment demonstrates that paperclip placement significantly impacts the flight distance of paper airplanes:

- The Middle Placement outperformed all other treatments, resulting in the longest flight distances. This placement likely optimizes stability and aerodynamic efficiency by creating a balanced center of gravity.
- The Rear Placement performed the worst, significantly underperforming compared to all other groups. A rear-heavy design destabilizes the airplane, reducing its aerodynamic effectiveness.
- The Nose Placement showed no significant difference from the Control group, suggesting that forward weight addition does not notably affect flight performance.
- Pairwise comparisons highlighted the significant differences between the Middle and Rear placements as well as the Rear and Nose placements, further validating the destabilizing effect of rear-heavy designs.

## Assumptions for model checking
```{r, echo=FALSE}
# Residual diagnostics
par(mfrow = c(2, 2))  
plot(model)  
```

- The Q-Q plot of residuals indicated that the residuals closely followed a straight line, confirming that the assumption of normality was not violated. 

- The residuals versus fitted values plot showed no clear patterns or trends, indicating that the variance of residuals remained constant across all levels of fitted values.

- The randomization of treatment order and the block design ensured that the observations for each treatment group were independent. 

Overall, the diagnostic plots (residual vs. fitted values, Q-Q plot, and scale-location plot) did not reveal any significant deviations or violations of model assumptions. This indicates that the ANOVA model is an appropriate choice for analyzing the impact of paperclip placement on flight performance.

# Discussion

## Summary of Results

This experiment investigated the impact of paperclip placement on the flight distance of paper airplanes, using a Randomized Complete Block Design (RCBD) to control for variability across airplanes. The analysis revealed that the placement of the paperclip significantly affects performance:

- Middle Placement:

  - Produced the longest flight distances, significantly outperforming all other placements.
  - The balanced center of gravity achieved with this placement likely optimizes stability and aerodynamic efficiency.
  
- Rear Placement:

  - Resulted in the shortest flight distances among all treatments, significantly underperforming the Middle, Nose, and Control groups.
  - This placement destabilizes the airplane by shifting the center of gravity backward, leading to the underperformance of the flight distance.

- Nose Placement vs. Control:

  - No significant difference was observed between the Nose and Control groups.
  - This suggests that forward weight placement has a minimal impact on flight performance compared to no weight addition.

The Middle placement consistently outperformed other configurations, while the Rear placement was the least effective. These results were confirmed through ANOVA and pairwise comparisons using Tukey's HSD test.

## Broader Interest

This experiment investigated how paperclip placement affects the flight distance of paper airplanes. The findings revealed that the middle placement produced the longest flight distances, significantly outperforming both the nose and rear placements. These results align with Scully et al. (2019), who demonstrated the importance of maintaining a balanced center of gravity for stable and efficient flight. A balanced weight distribution minimizes drag and optimizes lift, creating a more predictable and efficient flight path.

Smith and Johnson (2020) corroborate these findings, showing that balanced mass distribution optimizes stability and flight distance in paper airplanes. Their study emphasized that improper weight placement can introduce instability and unpredictable behavior during flight. The underperformance of rear-heavy designs observed in this study is consistent with Davis et al. (2018), who highlighted the destabilizing effects of an aft center of gravity on aerodynamic performance. They noted that shifting the center of gravity backward not only reduces stability but also increases drag, further impairing flight efficiency. The minimal improvement of the nose placement over the control group in this experiment also reflects the findings of Smith and Johnson (2020), who found that forward weight addition has limited benefits unless accompanied by other aerodynamic adjustments.

Overall, this experiment reinforces the critical role of weight distribution in flight mechanics. The significant advantage of the middle placement suggests that maintaining a well-balanced center of gravity enhances both stability and aerodynamic performance. These findings are directly applicable to real-world aviation practices, where precise weight distribution is essential. For instance, cargo distribution in commercial aircraft ensures stability and efficiency, while UAVs rely on accurate payload placement to maintain maneuverability and extend flight duration. The parallels between the experimental results and these real-world applications underscore the practical relevance of understanding and optimizing weight distribution in various aviation contexts.


## Limitations

- Sample size was limited to 30 airplanes, reducing statistical power. This can be addressed by increasing the number of airplanes to meet the calculated sample size requirement in future experiments. The limited sample size may have reduced the power to detect smaller effects.
- Measurement precision was limited to the nearest tenth decimal of an inch, potentially introducing rounding errors. Using a more precise measurement tool could help reduce this source of error.
- Variability in throwing technique may have influenced results, despite efforts to standardize the throwing motion. Implementing an automated machine could ensure consistent force and angle across all trials. Throwing inconsistencies may have added noise, potentially masking some treatment effects or exaggerating differences.
- Environmental factors, such as airflow from air conditioning (AC) systems, might have affected flight distances. Conducting trials in a fully controlled environment could further minimize variability.

# References

- Scully, J., et al. (2019). The impact of weight distribution on flight stability. Journal of Aerodynamics, 15(4), 123-134.

- Smith, J., & Johnson, L. (2020). The effect of mass distribution on the flight distance of paper airplanes. Journal of Experimental Aerodynamics, 22(3), 215-230.

- Davis, M., et al. (2018). Aerodynamic performance of paper planes with altered center of gravity. International Journal of Lightweight Structures and Aerodynamics, 10(2), 89-101.

\newpage
# Appendix

```{r, echo=FALSE}
print(results)
```
